Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging

### 2012-06-18 HRUIonGPU/jrnl_article.tex
Forgotten updates to uffc article. capon on speed

%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
\IEEEPARstart{C}{apon} adaptive beamforming applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. Although the method introduce some interesting trade-offs in addition to just improved resolution, like maintaining image resolution combined with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.  In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied on a typical cardiac data set. To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows. Earlier attempts on real time Capon beamforming have been focused around radar and passive sonar using intricate systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.  A systolic array processor is custom made hardware, and even if the computational power might be higher for selected problems compared with a GPU, the cost related to development and flexibility is much higher. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011} have previously implemented the Capon beamformer on a GPU, however their implementation was based on real data, where the Capon beamformer assumes complex. Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced compared with use of complex data. It is therefore crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows. All results in this paper is reported with use of critically sampled complex data, both laterally and in range.

In the next section we present background information about the Capon beamformer and the GPU computing model. In Section \ref{sec:meth} we go in detail about the parallel implementation of Capon beamforming, before we in Section \ref{sec:meth2} introduce methods to apply when the problem size increases. Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

\section{Background}
\subsection{Capon Beamforming}
When sensing the world around us with an array of elements, we perform a spatio-temporal filtering operation known as beamforming. The conventional form of beamforming is delay-and-sum (DAS), where the signal recorded on each element, $x_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
In conventional beamforming, the weighting $\vec{w} \in \mathbb{R}$ is usually applied to trade resolution for a better signal-to-noise ratio (SNR). Hence, by increasing the main lobe width of the system while gaining lower side lobe levels.

@@

The CUDA compute model is based on execution of a kernel-function across a grid of compute threads. Each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be easily synchronized if needed. 
%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
From a CUDA perspective the GPU consist of one or more stream multi processors (SM's), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions. A group of blocks is scheduled to each SM until all blocks have been processed. More important, the SM has a limited amount of near-core memory and registers available. Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency. Global read and writes on the GPU, in addition to CPU-GPU transfers have high penalty, and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.

>>>>

%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
\IEEEPARstart{C}{apon} adaptive beamforming applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. Although the method introduce some interesting trade-offs in addition to just improved resolution, like maintaining image resolution combined with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.  In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac setting. To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows. Earlier attempts on real time Capon beamforming have been focused around radar and passive sonar systems using intricate systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.  A systolic array processor is custom made hardware, and even if the computational throughput might be higher for selected problems compared with a GPU, the cost related to development and flexibility is considered much higher. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011} have previously implemented the Capon beamformer on a GPU, however their implementation was based on real data, where the Capon beamformer assumes complex. Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced using real data compared with use of complex data. It is crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows. All results in this paper is reported with use of critically sampled complex data, both laterally and in range.

In the next section we present background information about the Capon beamformer and the GPU computing model. In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:meth2} we introduce methods to apply as the problem size increases. Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

\section{Background}
\subsection{Capon Beamforming}
When sensing the world around us with an array of elements, we perform a spatio-temporal filtering operation known as beamforming. The conventional form of beamforming is delay-and-sum (DAS), where the signal recorded on each element, $x_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
In conventional beamforming, the weighting $\vec{w}$ is usually real, and applied to trade resolution for a better signal-to-noise ratio (SNR). Hence, by increasing the main lobe width of the system while gaining lower side lobe levels.

@@

The CUDA compute model is based on execution of a kernel-function across a grid of compute threads. Each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be easily synchronized if needed. 
%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
From a CUDA perspective the GPU consist of one or more stream multi processors (SM's), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions. A group of blocks is scheduled to each SM until all blocks have been processed. More important, the SM has a limited amount of near-core memory and registers available. Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency. Global read and writes on the GPU, in addition to CPU-GPU transfers have high penalty, and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration. In the next section, we will describe how this mapping should be done for Capon beamforming of ultrasound data.
